---
title: "gwverse: Case for Support"
fontsize: 11pt
geometry: "left=2cm,right=2cm,top=2cm,bottom=2cm"
output: 
  pdf_document:
    latex_engine: xelatex
subparagraph: yes
header-includes: |
  \usepackage{titlesec}
  \titlespacing{\title}{0pt}{\parskip}{-\parskip}
urlcolor: blue
linkcolor: black
mainfont: Arial
bibliography: gw_SI_paper.bib
csl: vancouver-brackets.csl
---

\vspace{-10truemm}

 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center',message=FALSE,warning=FALSE)
library(knitr)
library(kableExtra)
```


```{r echo = F, message = F, warning=F}
library(tidyverse)

df = read.csv("scopus_GW_GWR_GeogWeightReg_Eng.csv")
df %>% 
  filter(Year > "1995" & Year < 2022) %>% 
  group_by(Year) %>% 
  summarise(`Annual Publication Count` = n()) %>%
  ggplot(mapping = aes(x = Year, y = `Annual Publication Count`)) + 
  geom_bar(stat="identity") -> P1

```
## Overview 

This proposal describes a project that will drive the redevelopment of a well-used set of tools in spatial analysis based on Geographically Weighted (GW) frameworks, of which the best known is Geographically Weighted Regression (GWR)[@brunsdon1996geographically]. Our initial work has started this activity and created a new `gwverse` package (https://github.com/gwverse) which re-imagines current GW tools using a new modular framework and a new approach to function creation (ref to ArXiv doc TBC). We now want to extend this package development, to ensure the long-term sustainability of this heavily used open source research research software, through a community-led approach as described below.




## References
